---
title: "Autism Spectrum Disorder Methylation Analysis - A Look At Covariates"
author: "Namitha Mohandras, Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: https://github.com/markziemann/asd_meth

## Introduction

Here we are looking at the clinical covariates, to see the correlation structure between
ADOS, SRS, motor skills, language, IQ and anything else I can find.
There is no methylation data in this analysis.

```{r, packages}

suppressPackageStartupMessages({
  library("kableExtra")
  library("dplyr")
  library("RColorBrewer")
  library("gplots")
  library("ggplot2")
  library("beeswarm")
})

```

## Load data

Load the clinical characteristics.

```{r,load_data1}

x <- read.csv("ASD_EPIC_DATA/ASD_blood_summarysheet.csv")

```

## Exploratory analysis

Make an overview of the clinical data with a heatmap.

```{r,heat1}

dat <- x[,8:20]

heatmap.2(t(dat),trace="none",scale="row",mar=c(5,12))

heatmap.2(cor(dat,method="s"),trace="none",scale="none",mar=c(12,12),
  main="spearman correlation of covariates")

heatmap.2(cor(t(dat),method="s"),trace="none",scale="none",mar=c(12,12), 
  main="spearman correlation of participants")


```

## Dimensions of ASD

Let's take a look at the correlations of dimensions:

* Motor skills

* Social interaction score (ADOS)

* Diagnosis

* Social responsiveness score (SRS)

* Language

* IQ

```{r,dimensions}

head(dat)
dat2 <- dat[,8:13]

heatmap.2(cor(dat2,method="s"),trace="none",scale="none",
  mar=c(12,12),cexRow=1,cexCol=1)

cor(dat2,method="s") %>%
  kbl(caption = "Spearman correlation of") %>%
  kable_paper("hover", full_width = F)

```

## Independent-Dependent Variable Analysis

We will be exploring how the dependent variables associate with the dependent variables.

Independent variables: Sex, Age, Birth_mode, Birth_complications, gestational_age,
Birth_weight.g., Birth_order, Help_breathing_start.

Dependent vaiables: Motor_skills, Social_interaction_on_ADOS, Diagnosis, SRS_social_scores,
language, IQ.

Ideally I should be able to create a function to do this more or less automatically,
but the data are in different forms (categories, ordinals, vectors), so it might be complicated.

First I will look at sex (2 categories).

```{r,sex}

table(x$Sex)
length(which(x$Sex==1))
length(which(x$Sex==2))

message("sex-motorskills")
wilcox.test(subset(x,Sex==1)$Motor_skills,subset(x,Sex==2)$Motor_skills)
l <- list("sex1"=subset(x,Sex==1)$Motor_skills,"sex2"=subset(x,Sex==2)$Motor_skills)
beeswarm(l,ylab="motor skills",cex=2,pch=19)

message("sex-ados")
wilcox.test(subset(x,Sex==1)$Social_interaction_on_ADOS,subset(x,Sex==2)$Social_interaction_on_ADOS)
l <- list("sex1"=subset(x,Sex==1)$Social_interaction_on_ADOS,"sex2"=subset(x,Sex==2)$Social_interaction_on_ADOS)
beeswarm(l,ylab="ADOS",cex=2,pch=19)

message("sex-diagnosis")
wilcox.test(subset(x,Sex==1)$Diagnosis,subset(x,Sex==2)$Diagnosis)
l <- list("sex1"=subset(x,Sex==1)$Diagnosis,"sex2"=subset(x,Sex==2)$Diagnosis)
beeswarm(l,ylab="Diagnosis",cex=2,pch=19)

message("sex-srs")
wilcox.test(subset(x,Sex==1)$SRS_social_scores,subset(x,Sex==2)$SRS_social_scores)
l <- list("sex1"=subset(x,Sex==1)$SRS_social_scores,"sex2"=subset(x,Sex==2)$SRS_social_scores)
beeswarm(l,ylab="SRS",cex=2,pch=19)

message("sex-language")
wilcox.test(subset(x,Sex==1)$language,subset(x,Sex==2)$language)
l <- list("sex1"=subset(x,Sex==1)$language,"sex2"=subset(x,Sex==2)$language)
beeswarm(l,ylab="language",cex=2,pch=19)

message("sex-IQ")
wilcox.test(subset(x,Sex==1)$IQ,subset(x,Sex==2)$IQ)
l <- list("sex1"=subset(x,Sex==1)$IQ,"sex2"=subset(x,Sex==2)$IQ)
beeswarm(l,ylab="IQ",cex=2,pch=19)

message("sex-famID")
wilcox.test(subset(x,Sex==1)$Family_ID,subset(x,Sex==2)$Family_ID)
l <- list("sex1"=subset(x,Sex==1)$Family_ID,"sex2"=subset(x,Sex==2)$Family_ID)
beeswarm(l,ylab="Family_ID",cex=2,pch=19)

```

Now I will look at Birth_mode (categorical).

```{r,birthmode}

table(x$Birth_mode)

message("birthmode-motorskills")
wilcox.test(subset(x,Birth_mode==1)$Motor_skills,subset(x,Birth_mode==2)$Motor_skills)
l <- list("Birth_mode1"=subset(x,Birth_mode==1)$Motor_skills,"Birth_mode2"=subset(x,Birth_mode==2)$Motor_skills)
beeswarm(l,ylab="motor skills",cex=2,pch=19)

message("Birth_mode-ados")
wilcox.test(subset(x,Birth_mode==1)$Social_interaction_on_ADOS,subset(x,Birth_mode==2)$Social_interaction_on_ADOS)
l <- list("Birth_mode1"=subset(x,Birth_mode==1)$Social_interaction_on_ADOS,"Birth_mode2"=subset(x,Birth_mode==2)$Social_interaction_on_ADOS)
beeswarm(l,ylab="ADOS",cex=2,pch=19)

message("Birth_mode-diagnosis")
wilcox.test(subset(x,Birth_mode==1)$Diagnosis,subset(x,Birth_mode==2)$Diagnosis)
l <- list("Birth_mode1"=subset(x,Birth_mode==1)$Diagnosis,"Birth_mode2"=subset(x,Birth_mode==2)$Diagnosis)
beeswarm(l,ylab="Diagnosis",cex=2,pch=19)

message("Birth_mode-srs")
wilcox.test(subset(x,Birth_mode==1)$SRS_social_scores,subset(x,Birth_mode==2)$SRS_social_scores)
l <- list("Birth_mode1"=subset(x,Birth_mode==1)$SRS_social_scores,"Birth_mode2"=subset(x,Birth_mode==2)$SRS_social_scores)
beeswarm(l,ylab="SRS",cex=2,pch=19)

message("Birth_mode-language")
wilcox.test(subset(x,Birth_mode==1)$language,subset(x,Birth_mode==2)$language)
l <- list("Birth_mode1"=subset(x,Birth_mode==1)$language,"Birth_mode2"=subset(x,Birth_mode==2)$language)
beeswarm(l,ylab="language",cex=2,pch=19)

message("Birth_mode-IQ")
wilcox.test(subset(x,Birth_mode==1)$IQ,subset(x,Birth_mode==2)$IQ)
l <- list("Birth_mode1"=subset(x,Birth_mode==1)$IQ,"Birth_mode2"=subset(x,Birth_mode==2)$IQ)
beeswarm(l,ylab="IQ",cex=2,pch=19)

message("Birth_mode-famID")
wilcox.test(subset(x,Birth_mode==1)$Family_ID,subset(x,Birth_mode==2)$Family_ID)
l <- list("Birth_mode1"=subset(x,Birth_mode==1)$Family_ID,"Birth_mode2"=subset(x,Birth_mode==2)$Family_ID)
beeswarm(l,ylab="Family_ID",cex=2,pch=19)

```

Now look at breathing assistance (categorical).

```{r,breathing}

message("Help_breathing_start-motorskills")
wilcox.test(subset(x,Help_breathing_start==0)$Motor_skills,subset(x,Help_breathing_start==1)$Motor_skills)
l <- list("Help_breathing_start1"=subset(x,Help_breathing_start==0)$Motor_skills,"Help_breathing_start2"=subset(x,Help_breathing_start==1)$Motor_skills)
beeswarm(l,ylab="motor skills",cex=2,pch=19)

message("Help_breathing_start-ados")
wilcox.test(subset(x,Help_breathing_start==0)$Social_interaction_on_ADOS,subset(x,Help_breathing_start==1)$Social_interaction_on_ADOS)
l <- list("Help_breathing_start1"=subset(x,Help_breathing_start==0)$Social_interaction_on_ADOS,"Help_breathing_start2"=subset(x,Help_breathing_start==1)$Social_interaction_on_ADOS)
beeswarm(l,ylab="ADOS",cex=2,pch=19)

message("Help_breathing_start-diagnosis")
wilcox.test(subset(x,Help_breathing_start==0)$Diagnosis,subset(x,Help_breathing_start==1)$Diagnosis)
l <- list("Help_breathing_start1"=subset(x,Help_breathing_start==0)$Diagnosis,"Help_breathing_start2"=subset(x,Help_breathing_start==1)$Diagnosis)
beeswarm(l,ylab="Diagnosis",cex=2,pch=19)

message("Help_breathing_start-srs")
wilcox.test(subset(x,Help_breathing_start==0)$SRS_social_scores,subset(x,Help_breathing_start==1)$SRS_social_scores)
l <- list("Help_breathing_start1"=subset(x,Help_breathing_start==0)$SRS_social_scores,"Help_breathing_start2"=subset(x,Help_breathing_start==1)$SRS_social_scores)
beeswarm(l,ylab="SRS",cex=2,pch=19)

message("Help_breathing_start-language")
wilcox.test(subset(x,Help_breathing_start==0)$language,subset(x,Help_breathing_start==1)$language)
l <- list("Help_breathing_start1"=subset(x,Help_breathing_start==0)$language,"Help_breathing_start2"=subset(x,Help_breathing_start==1)$language)
beeswarm(l,ylab="language",cex=2,pch=19)

message("Help_breathing_start-IQ")
wilcox.test(subset(x,Help_breathing_start==0)$IQ,subset(x,Help_breathing_start==1)$IQ)
l <- list("Help_breathing_start1"=subset(x,Help_breathing_start==0)$IQ,"Help_breathing_start2"=subset(x,Help_breathing_start==1)$IQ)
beeswarm(l,ylab="IQ",cex=2,pch=19)

message("Help_breathing_start-famID")
wilcox.test(subset(x,Help_breathing_start==0)$Family_ID,subset(x,Help_breathing_start==1)$Family_ID)
l <- list("Help_breathing_start1"=subset(x,Help_breathing_start==0)$Family_ID,"Help_breathing_start2"=subset(x,Help_breathing_start==1)$Family_ID)
beeswarm(l,ylab="Family_ID",cex=2,pch=19)

```

Now look at birth complications (categorical).

```{r,birthcomplications}

table(x$Birth_complications)

message("birthmode-motorskills")
wilcox.test(subset(x,Birth_complications==0)$Motor_skills,subset(x,Birth_complications>0)$Motor_skills)
l <- list("Birth_complications1"=subset(x,Birth_complications==0)$Motor_skills,"Birth_complications2"=subset(x,Birth_complications>0)$Motor_skills)
beeswarm(l,ylab="motor skills",cex=2,pch=19)

message("Birth_complications-ados")
wilcox.test(subset(x,Birth_complications==0)$Social_interaction_on_ADOS,subset(x,Birth_complications>0)$Social_interaction_on_ADOS)
l <- list("No_Birth_complications"=subset(x,Birth_complications==0)$Social_interaction_on_ADOS,"Birth_complications"=subset(x,Birth_complications>0)$Social_interaction_on_ADOS)
beeswarm(l,ylab="ADOS",cex=2,pch=19)

message("Birth_complications-diagnosis")
wilcox.test(subset(x,Birth_complications==0)$Diagnosis,subset(x,Birth_complications>0)$Diagnosis)
l <- list("No_Birth_complications"=subset(x,Birth_complications==0)$Diagnosis,"Birth_complications"=subset(x,Birth_complications>0)$Diagnosis)
beeswarm(l,ylab="Diagnosis",cex=2,pch=19)

message("Birth_complications-srs")
wilcox.test(subset(x,Birth_complications==0)$SRS_social_scores,subset(x,Birth_complications>0)$SRS_social_scores)
l <- list("No_Birth_complications"=subset(x,Birth_complications==0)$SRS_social_scores,"Birth_complications"=subset(x,Birth_complications>0)$SRS_social_scores)
beeswarm(l,ylab="SRS",cex=2,pch=19)

message("Birth_complications-language")
wilcox.test(subset(x,Birth_complications==0)$language,subset(x,Birth_complications>0)$language)
l <- list("No_Birth_complications"=subset(x,Birth_complications==0)$language,"Birth_complications"=subset(x,Birth_complications>0)$language)
beeswarm(l,ylab="language",cex=2,pch=19)

message("Birth_complications-IQ")
wilcox.test(subset(x,Birth_complications==0)$IQ,subset(x,Birth_complications>0)$IQ)
l <- list("No_Birth_complications"=subset(x,Birth_complications==0)$IQ,"Birth_complications"=subset(x,Birth_complications>0)$IQ)
beeswarm(l,ylab="IQ",cex=2,pch=19)

message("Birth_complications-famID")
wilcox.test(subset(x,Birth_complications==0)$Family_ID,subset(x,Birth_complications>0)$Family_ID)
l <- list("No_Birth_complications"=subset(x,Birth_complications==0)$Family_ID,"Birth_complications"=subset(x,Birth_complications>0)$Family_ID)
beeswarm(l,ylab="Family_ID",cex=2,pch=19)

```

Now look at gestational age (continuous).

```{r,gest_age}

x$gestational_age
summary(x$gestational_age)
hist(x$gestational_age)

message("gestational_age-motor_skills")
mylm <- lm(x$Motor_skills ~ x$gestational_age)
summary(mylm)
plot(x$gestational_age,x$Motor_skills,cex=2,pch=19)
abline(mylm,col="red",lty=2,lwd=2)

message("gestational_age-ADOS")
mylm <- lm(x$Social_interaction_on_ADOS ~ x$gestational_age)
summary(mylm)
plot(x$gestational_age,x$Social_interaction_on_ADOS,cex=2,pch=19)
abline(mylm,col="red",lty=2,lwd=2)

message("gestational_age-Diagnosis")
mylm <- lm(x$Diagnosis ~ x$gestational_age)
summary(mylm)
plot(x$gestational_age,x$Diagnosis,cex=2,pch=19)
abline(mylm,col="red",lty=2,lwd=2)

message("gestational_age-SRS")
mylm <- lm(x$SRS_social_scores ~ x$gestational_age)
summary(mylm)
plot(x$gestational_age,x$SRS_social_scores,cex=2,pch=19)
abline(mylm,col="red",lty=2,lwd=2)

message("gestational_age-language")
mylm <- lm(x$language ~ x$gestational_age)
summary(mylm)
plot(x$gestational_age,x$language,cex=2,pch=19)
abline(mylm,col="red",lty=2,lwd=2)

message("gestational_age-IQ")
mylm <- lm(x$IQ ~ x$gestational_age)
summary(mylm)
plot(x$gestational_age,x$IQ,cex=2,pch=19)
abline(mylm,col="red",lty=2,lwd=2)

```

Now look at birthweight (continuous).

```{r,birthweight}

x$Birth_weight.g.
summary(x$Birth_weight.g.)
hist(x$Birth_weight.g.)

message("birthweight-motor_skills")
mylm <- lm(x$Motor_skills ~ x$Birth_weight.g.)
summary(mylm)
plot(x$Birth_weight.g.,x$Motor_skills,cex=2,pch=19)
abline(mylm,col="red",lty=2,lwd=2)

message("Birth_weight.g.-ADOS")
mylm <- lm(x$Social_interaction_on_ADOS ~ x$Birth_weight.g.)
summary(mylm)
plot(x$Birth_weight.g.,x$Social_interaction_on_ADOS,cex=2,pch=19)
abline(mylm,col="red",lty=2,lwd=2)

message("Birth_weight.g.-Diagnosis")
mylm <- lm(x$Diagnosis ~ x$Birth_weight.g.)
summary(mylm)
plot(x$Birth_weight.g.,x$Diagnosis,cex=2,pch=19)
abline(mylm,col="red",lty=2,lwd=2)

message("Birth_weight.g.-SRS")
mylm <- lm(x$SRS_social_scores ~ x$Birth_weight.g.)
summary(mylm)
plot(x$Birth_weight.g.,x$SRS_social_scores,cex=2,pch=19)
abline(mylm,col="red",lty=2,lwd=2)

message("Birth_weight.g.-language")
mylm <- lm(x$language ~ x$Birth_weight.g.)
summary(mylm)
plot(x$Birth_weight.g.,x$language,cex=2,pch=19)
abline(mylm,col="red",lty=2,lwd=2)

message("Birth_weight.g.-IQ")
mylm <- lm(x$IQ ~ x$Birth_weight.g.)
summary(mylm)
plot(x$Birth_weight.g.,x$IQ,cex=2,pch=19)
abline(mylm,col="red",lty=2,lwd=2)

```


## Session information

```{r,sessioninfo}

sessionInfo()

```
