---
title: "Autism Spectrum Disorder Methylation analysis - comparison of Guthrie card and fresh blood"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: https://github.com/markziemann/asd_meth

## Introduction

Here I will be running a comparison of differential methylation data from guthrie and fresh blood samples.

Here are the files that I'm using:

* ASD_blood_top_dmps_onADOS_withintw_Nov27_limma.csv

* ASD_guthrie_top_dmps_onADOS_withintw_Nov27_limma.csv


```{r,libs}

suppressPackageStartupMessages({
  library("parallel")
  library("mitch")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  source("https://raw.githubusercontent.com/markziemann/gmea/main/meth_functions.R")
  library("data.table")
  library("kableExtra")
  library("eulerr")
  library("tictoc")
})

```


## Load data

```{r,load1}

bl_design <- readRDS(file="bl_design.rds")
bl_mvals <- readRDS(file="bl_mvals.rds")

gu_design <- readRDS(file="gu_design.rds")
gu_mvals <- readRDS(file="gu_mvals.rds")

```

## Probe sets

For each gene, extract out the probes.

```{r,probesets}

anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)
promoters <- myann[promoters,]
gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
sets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))
summary(unlist(lapply(sets,length)))

```

## GMEA functions

This is an enrichment technique.

```{r,func}

# get the median, mean and 1-sample t-test result for each gene
pmea <- function(mval,design,sets,cores=2) {
  fit <- lmFit(mval, design)
  fit <- eBayes(fit)
  top <- topTable(fit,coef=ncol(design),num=Inf, sort.by = "P")
  l <- mclapply(seq(1,length(sets)), function(i) {
    g <- names(sets[i])
    tstats <- top[rownames(top) %in% sets[[i]],"t"]
    myn <- length(tstats)
    mymean <- mean(tstats)
    mymedian <- median(tstats)
    if ( length(tstats) < 2 ) {
      pval=1
    } else {
      wtselfcont <- t.test(tstats)
      pval=wtselfcont$p.value
    }
    c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
      "P.Value"=pval)
  } , mc.cores=cores)

  df <- do.call(rbind, l)
  rownames(df) <- df[,1]
  df <- df[,-1]
  tmp <- apply(df,2,as.numeric)
  rownames(tmp) <- rownames(df)
  df <- as.data.frame(tmp)
  df$sig <- -log10(df[,4])
  df <- df[order(-df$sig),]
  df$FDR <- p.adjust(df$P.Value)
  out <- list("df"=df,"toptable"=top)
  return(out)
}
# pmea_res <- pmea(mval=mval,design=design,sets=head(sets,20),cores=detectCores()/2)

# Run the fry test for each gene. This is a more conservative test.
run_fry <- function(mval,design,sets,cores=2) {
  split_sets <- split(sets, ceiling(seq_along(sets)/200))
  fry_l <- mclapply(split_sets,function(l) {
    fry(y=mval, index = l, design = design,
      contrast = ncol(design) )
  } , mc.cores=cores )
  fry_res <- do.call(rbind,fry_l)
  rownames(fry_res) <- sub("\\.","@",rownames(fry_res))
  rownames(fry_res) <- sapply(strsplit(rownames(fry_res),"@"),"[[",2)
  fry_res[is.na(fry_res$PValue),"PValue"] <- 1
  fry_res <- fry_res[order(fry_res$PValue),]
  fry_res$FDR <- p.adjust(fry_res$PValue,method="fdr")
  return(fry_res)
}
#fry_res <- run_fry(mval=mval,design=design,sets=sets,cores=cores)

# main function to perform 1-sample t-test and fry test and merge the results.
main <- function(mval,design,sets,cores=2){
  pmea_res <- pmea(mval=mval,design=design,sets=sets,cores=cores)
  pmea_df <- pmea_res[[1]]
  limma_df <- pmea_res[[2]]
  fry_res <- run_fry(mval=mval,design=design,sets=sets,cores=cores)
  m <- merge(pmea_df,fry_res,by=0)
  rownames(m) <- m$Row.names
  m$Row.names = NULL
  m <- m[,c("nprobes","median","PValue","FDR.y")]
  colnames(m) <- c("nprobes","median","PValue","FDR")
  m <- m[order(m$PValue),]
  out <- list("res"=m,"limma_df"=limma_df)
  return(out)
}
#res <- main(mval,design,sets,cores=detectCores()/2)

probe_bias <- function(res) {
  res$sig <- -log10(res$PValue)
  sig <- subset(res,FDR < 0.05)
  plot(res$nprobes,res$sig,log="x",
    pch=19,cex=0.6,
    xlab="no. probes",ylab="-log10(p-value)")
  points(sig$nprobes,sig$sig,col="red",pch=19,cex=0.62)
  SIG = nrow(sig)
  TOT = nrow(res)
  HEADER <- paste(TOT, "genes in total.", SIG, "with FDR<0.05.")
  mtext(HEADER)
}

volcano_plot <- function(res) {
  res$sig <- -log10(res$PValue)
  sig <- subset(res,FDR < 0.05)
  plot(res$median,res$sig,
    pch=19,cex=0.6,
    xlab="median t-statistic",ylab="-log10(p-value)")
  points(sig$median,sig$sig,col="red",pch=19,cex=0.62)
  SIG = nrow(sig)
  UP = nrow(subset(sig,median>0))
  DN = nrow(subset(sig,median<0))
  TOT = nrow(res)
  HEADER <- paste(TOT, "genes in total.", SIG, "with FDR<0.05;",DN,"down,",UP,"up")
  mtext(HEADER)
}

gmea_boxplot <- function(res,sets,n=50) {
  df <- res[[1]]
  limma_df <- res[[2]]
  # smallest pval
  par(mfrow=c(1,2))
  gs <- head(rownames(df),n)
  mysets <- sets[names(sets) %in% gs]
  tstats <- lapply(mysets, function(set) {
    limma_df[rownames(limma_df) %in% set,"t"]
  })
  tstats <- tstats[order(unlist(lapply(tstats,median)))]
  boxplot(tstats,horizontal=TRUE,las=1,
    main="smallest p-val",cex.axis=0.6,
    xlab="t-statistic")
  grid()
  # biggest effect size (median)
  sig <- subset(df,FDR < 0.05)
  gs <- head(rownames(sig[order(-abs(sig$median)),]),n)
  if ( length(gs) >2 ) {
    tstats <- lapply(gs, function(g) {
      df[which(df$genes==g),"tvals"]
    })
    names(tstats) <- gs
    tstats <- tstats[order(unlist(lapply(tstats,median)))]
    boxplot(tstats,horizontal=TRUE,las=1,
      main="biggest effect size(median)",cex.axis=0.6,
      xlab="t-statistic")
    grid()
  } else {
    plot(1)
    mtext("too few significant genes found")
  }
    par(mfrow=c(1,1))
}

```

## GMEA for blood

Effect size is estimated with the median t-statistic for each gene.
Significance is estimated using the fry test.
Results are merged for the user.

* PMEA step 196.063 sec elapsed on 8 cores

* Fry step 293.569 sec elapsed on 8 cores

* Fry single core 1205.519 sec

```{r,bl1}

bl_design <- readRDS(file="bl_design.rds")
bl_mvals <- readRDS(file="bl_mvals.rds")

tic()
res_bl <- main(mval=bl_mvals,design=bl_design,sets,cores=detectCores()/2)
res_bl_df <- res_bl[[1]] # enrichment results
res_bl_top <- res_bl[[2]] # limma results
toc() ## 502-522 sec elapsed on 8 cores (8.3-8.7 mins)

head(res_bl_df,30) %>%
  kbl(caption="Top DM genes in blood identified with GMEA") %>%
  kable_paper("hover", full_width = F)

```

Now some visualisations.

```{r,bl2}

volcano_plot(res_bl_df)
probe_bias(res_bl_df)
gmea_boxplot(res_bl,sets)

```

##  GMEA for guthrie

```{r,gu1}

gu_design <- readRDS(file="gu_design.rds")
gu_mvals <- readRDS(file="gu_mvals.rds")

tic()
res_gu <- main(mval=gu_mvals,design=gu_design,sets,cores=detectCores()/2)
res_gu_df <- res_gu[[1]] # enrichment results
res_gu_top <- res_gu[[2]] # limma results
toc() ## 502-522 sec elapsed on 8 cores (8.3-8.7 mins)

head(res_gu_df,30) %>%
  kbl(caption="Top DM genes in Guthrie card identified with GMEA") %>%
  kable_paper("hover", full_width = F)

```

Now some visualisations.

```{r,gu2}

volcano_plot(res_gu_df)
probe_bias(res_gu_df)
gmea_boxplot(res_gu,sets)

```

## Comparison of blood and guthrie card


```{r,compare1}

res_bl_df$bl <- res_bl_df$median
res_gu_df$gu <- res_gu_df$median

m1 <- merge(res_bl_df,res_gu_df,by=0)

plot(m1$bl,m1$gu,xlab="bl",ylab="gu") ; grid()
plot(rank(m1$bl),rank(m1$gu),xlab="bl rank",ylab="gu rank",cex=0.6,pch=19) ; grid()

```

## Mitch enrichment analysis

```{r,genesets}

genesets <- gmt_import("ReactomePathways.gmt")

```

Here I'm using the median t-statistic for downstream enrichment.
Pathways are from REACTOME and analysis using mitch.

```{r,mitch_direction,eval=FALSE}

#res_bl_df$bl <- sign(res_bl_df$median) * -log10(res_bl_df$PValue)
#res_gu_df$gu <- sign(res_gu_df$median) * -log10(res_gu_df$PValue)

#res_bl_df$bl <- res_bl_df$median * -log10(res_bl_df$PValue)
#res_gu_df$gu <- res_gu_df$median * -log10(res_gu_df$PValue)

res_bl_df$bl <- res_bl_df$median
res_gu_df$gu <- res_gu_df$median

bl <-  res_bl_df[,"bl",drop=FALSE]
mitch_bl <- mitch_calc(bl,genesets,priority="effect")
head( mitch_bl$enrichment_result,30) %>%
  kbl(caption="BLOOD: Top effect size pathways found with mitch") %>%
  kable_paper("hover", full_width = F)

sig <- subset(mitch_bl$enrichment_result,`p.adjustANOVA`<0.05)
head(sig[order(-abs(sig$s.dist)),],30) %>%
  kbl(caption="BLOOD: Top effect size pathways found with mitch after 1% FDR filtering") %>%
  kable_paper("hover", full_width = F)

gu <-  res_gu_df[,"gu",drop=FALSE]
mitch_gu <- mitch_calc(gu,genesets,priority="effect")
head( mitch_gu$enrichment_result,30) %>%
  kbl(caption="GUTHRIE: Top effect size pathways found with mitch") %>%
  kable_paper("hover", full_width = F)

sig <- subset(mitch_gu$enrichment_result,`p.adjustANOVA`<0.05)
head(sig[order(-abs(sig$s.dist)),],30) %>%
  kbl(caption="GUTHRIE: Top effect size pathways found with mitch after 1% FDR filtering") %>%
  kable_paper("hover", full_width = F)

```

Mitch joint analysis of blood and guthrie card.

As using the median value is apparently the most sensitive, we'll go with that.

```{r,mitch_joint}

m <- merge(res_bl_df,res_gu_df,by=0)
rownames(m) <- m$Row.names
m2 <- m[,c("bl","gu")]
m2[is.na(m2)] <- 0

res <- mitch_calc(m2, genesets, priority="effect")

mitch_report(res,"blgu_mitch.html",overwrite=TRUE)

sig <- subset(res$enrichment_result,p.adjustMANOVA<0.01)

head(sig[order(-abs(sig$s.dist)),],30) %>%
  kbl(caption="Top pathways found with mitch (effect size after 1% FDR filtering)") %>%
  kable_paper("hover", full_width = F)

```

## Camera enrichment analysis

Camera is designed to control for correlation bias.

Using the median value gives better results than the directional significance or the combined metric.

Blood dataset.

Need to filter by FDR then sort by effect size.

```{r,cam_bl}

stat <- res_bl_df$median
#stat <- res_bl_df$median * -log10(res_bl_df$PValue)
#stat <- sign(res_bl_df$median) * -log10(res_bl_df$PValue)

names(stat) <- rownames(res_bl_df)
stat[is.na(stat)] <- 0

tic()
cres <- cameraPR(statistic=stat, index=genesets, use.ranks = FALSE, inter.gene.cor=0.01, sort = TRUE)
cres <- subset(cres,NGenes>4)
cres$FDR <- p.adjust(cres$PValue,method="fdr")
mymedians <- sapply(1:nrow(cres),function(i) {
  myset <- genesets[[which(names(genesets) == rownames(cres)[i])]]
  mystats <- stat[names(stat) %in% myset]
  median(mystats)
})
cres$median <- mymedians
toc() # 1.0 sec

sig <- subset(cres,FDR<0.05)

head(sig[order(-abs(sig$median)),],20) %>%
  kbl(caption = "Top effect size pathways found with CAMERA after 5% FDR filtering") %>%
  kable_paper("hover", full_width = F)

```

Now drilling down to the top set.

```{r,cam_bl_topset}

myset <- rownames(head(sig[order(-abs(sig$median)),],1))
myset_members <- genesets[[which(names(genesets) == myset)]]
mystats <- stat[which(names(stat) %in% myset_members)]
myl <- list("allgenes"=stat,myset=mystats)
boxplot(myl,cex=0)

library(vioplot)
vioplot(myl,main=myset)

head(mystats[order(mystats)],20)

```

Guthrie dataset.

```{r,cam_guth}

stat <- res_gu_df$median
#stat <- res_gu_df$median * -log10(res_gu_df$PValue)
#stat <- sign(res_gu_df$median) * -log10(res_gu_df$PValue)

names(stat) <- rownames(res_gu_df)
stat[is.na(stat)] <- 0

tic()
cres <- cameraPR(statistic=stat, index=genesets, use.ranks = FALSE, inter.gene.cor=0.01, sort = TRUE)
cres <- subset(cres,NGenes>4)
cres$FDR <- p.adjust(cres$PValue,method="fdr")
mymedians <- sapply(1:nrow(cres),function(i) {
  myset <- genesets[[which(names(genesets) == rownames(cres)[i])]]
  mystats <- stat[names(stat) %in% myset]
  median(mystats)
})
cres$median <- mymedians
toc() # 1.0 sec

sig <- subset(cres,FDR<0.05)

head(sig[order(-abs(sig$median)),],20) %>%
  kbl(caption = "Top effect size pathways found with CAMERA after 5% FDR filtering") %>%
  kable_paper("hover", full_width = F)

```


## Session information

For reproducibility

```{r,session}

sessionInfo()

```
