---
title: "Correlates of diagnosis and methylation - compartment analysis"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: https://github.com/markziemann/asd_meth

## Introduction

Here I will be running a comparison of differential methylation data from guthrie and fresh blood samples.

Here are the files that I'm using:

* limma_guthrie_diagnosis.csv

* limma_blood_diagnosis.csv

```{r,libs}

suppressPackageStartupMessages({
  library("parallel")
  library("mitch")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  source("https://raw.githubusercontent.com/markziemann/gmea/main/meth_functions.R")
  library("data.table")
  library("kableExtra")
  library("eulerr")
  library("RIdeogram")
  library("GenomicRanges")
  library("tictoc")
})

source("meth_functions.R")

```


## Load Limma and RUV data for manhattan plot

First I will generate plots for limmma analysis.

```{r,manhattan1}

# blood at assessment
top <- read.csv("limma_blood_diagnosis.csv")
nrow(top)
top <- subset(top,P.Value<1e-2)
nrow(top)
-log10(min(top$P.Value))
top$chr <- as.integer(gsub("chr","",top$chr))
top$snp <- paste(top$Name,top$UCSC_RefGene_Name)
top$snp <- sapply(strsplit(top$snp,";"),"[[",1)
up <- subset(top,logFC>0)
dn <- subset(top,logFC<0)

if ( (head(up$P.Value,1) < 1e-4) & (head(dn$P.Value,1) < 1e-4 ) ) {
  par(mfrow=c(2,1))
  par(mar=c(4,4,3,5))
  manhattan(x=up,chr="chr",bp="pos",p="P.Value",snp="snp",suggestiveline = -log10(1e-04),
    ylim=c(2,6), annotatePval= 1e-04 ,main="Blood at assessment limma hypermethylated")
  manhattan(x=dn,chr="chr",bp="pos",p="P.Value",snp="snp",suggestiveline = -log10(1e-04),
    ylim=c(2,6), annotatePval= 1e-04 , main="Blood at assessment limma hypomethylated")
  pdf("manhat_limma_bl.pdf")
  par(mfrow=c(2,1))
  par(mar=c(4,4,3,5))
  manhattan(x=up,chr="chr",bp="pos",p="P.Value",snp="snp",suggestiveline = -log10(1e-04),
    ylim=c(2,6), annotatePval= 1e-04 ,main="Blood at assessment limma hypermethylated")
  manhattan(x=dn,chr="chr",bp="pos",p="P.Value",snp="snp",suggestiveline = -log10(1e-04),
    ylim=c(2,6), annotatePval= 1e-04 , main="Blood at assessment limma hypomethylated")
  dev.off()
}

# neonatal guthrie card
top <- read.csv("limma_guthrie_diagnosis.csv")
nrow(top)
top <- subset(top,P.Value<1e-2)
nrow(top)
-log10(min(top$P.Value))
top$chr <- as.integer(gsub("chr","",top$chr))
top$snp <- paste(top$Name,top$UCSC_RefGene_Name)
top$snp <- sapply(strsplit(top$snp,";"),"[[",1)
up <- subset(top,logFC>0)
dn <- subset(top,logFC<0)

if ( (head(up$P.Value,1) < 1e-4) & (head(dn$P.Value,1) < 1e-4 ) ) {
  par(mfrow=c(2,1))
  par(mar=c(4,4,3,5))
  manhattan(x=up,chr="chr",bp="pos",p="P.Value",snp="snp",suggestiveline = -log10(1e-04),
    ylim=c(2,6), annotatePval= 1e-04 ,main="Neonatal Guthrie card limma hypermethylated",
    annotateTop=FALSE)
  manhattan(x=dn,chr="chr",bp="pos",p="P.Value",snp="snp",suggestiveline = -log10(1e-04),
    ylim=c(2,6), annotatePval= 1e-04 , main="Neonatal Guthrie card limma hypomethylated",
    annotateTop=FALSE)
  pdf("manhat_limma_gu.pdf")
  par(mfrow=c(2,1))
  par(mar=c(4,4,3,5))
  manhattan(x=up,chr="chr",bp="pos",p="P.Value",snp="snp",suggestiveline = -log10(1e-04),
    ylim=c(2,6), annotatePval= 1e-04 ,main="Neonatal Guthrie card limma hypermethylated",
    annotateTop=FALSE)
  manhattan(x=dn,chr="chr",bp="pos",p="P.Value",snp="snp",suggestiveline = -log10(1e-04),
    ylim=c(2,6), annotatePval= 1e-04 , main="Neonatal Guthrie card limma hypomethylated",
    annotateTop=FALSE)
  dev.off()
}

```

## Compartment enrichment

Looking for enrichments in different genomic compartments.

Guthrie cards.

```{r,comp1a}

par(mfrow=c(2,1))

# guthrie
dma1 <- read.csv("limma_guthrie_diagnosis.csv")
if (nrow(subset(dma1,P.Value <1e-4)) > 100 ) {
comp <- compartment_enrichment(dma1)
comp <- lapply(comp,function(x) {subset(x,OR!=0)} )
comp
make_forest_plots_up(comp)
make_forest_plots_dn(comp)
}

```

Blood at assessment.

```{r,comp1b}

# blood
dma2 <- read.csv("limma_blood_diagnosis.csv")
if (nrow(subset(dma2,P.Value <1e-4)) > 100 ) {
comp <- compartment_enrichment(dma2)
comp <- lapply(comp,function(x) {subset(x,OR!=0)} )
comp
make_forest_plots_up(comp)
make_forest_plots_dn(comp)
}

```

## Compartments top 1000

Guthrie cards.

```{r,comp2a}

par(mfrow=c(2,1))
# guthrie
comp <- compartment_enrichment2(dma1)
comp <- lapply(comp,function(x) {subset(x,OR!=0)} )
comp
make_forest_plots_up(comp)
make_forest_plots_dn(comp)

```
Blood at assessment.

```{r,comp2b}

# blood
comp <- compartment_enrichment2(dma2)
comp <- lapply(comp,function(x) {subset(x,OR!=0)} )
comp
make_forest_plots_up(comp)
make_forest_plots_dn(comp)

```

## Looking for enrichments in proximity to CGI contexts

Guthrie cards.

```{r,cgi1a}

par(mfrow=c(2,1))

if (nrow(subset(dma1,P.Value <1e-4)) > 100 ) {
# guthrie
comp <- cgi_enrichment(dma1)
comp <- lapply(comp,function(x) {subset(x,OR!=0)} )
comp
make_forest_plots_up(comp)
make_forest_plots_dn(comp)
}

```

Blood at assessment.

```{r,cgi1b}

if (nrow(subset(dma2,P.Value <1e-4)) > 100 ) {
# blood
comp <- cgi_enrichment(dma2)
comp <- lapply(comp,function(x) {subset(x,OR!=0)} )
comp
make_forest_plots_up(comp)
make_forest_plots_dn(comp)
}

```

## CGI enrichments in top 1000 probes

Guthrie cards.

```{r,cgi2a}

par(mfrow=c(2,1))
# guthrie
comp <- cgi_enrichment2(dma1)
comp <- lapply(comp,function(x) {subset(x,OR!=0)} )
comp
make_forest_plots_up(comp)
make_forest_plots_dn(comp)

```

Blood at assessment.

```{r,cgi2b}

# blood
comp <- cgi_enrichment2(dma2)
comp <- lapply(comp,function(x) {subset(x,OR!=0)} )
comp
make_forest_plots_up(comp)
make_forest_plots_dn(comp)

```

## Session information

For reproducibility

```{r,session}

sessionInfo()

```
