---
title: "Correlates of motor impairment and methylation - enrichment analysis"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: https://github.com/markziemann/asd_meth

## Introduction

Here I will be running a comparison of differential methylation data from guthrie and fresh blood samples.

Here are the files that I'm using:

* limma_blood_motor.csv

* limma_guthrie_motor.csv

```{r,libs}

suppressPackageStartupMessages({
  library("parallel")
  library("mitch")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  source("https://raw.githubusercontent.com/markziemann/gmea/main/meth_functions.R")
  library("data.table")
  library("kableExtra")
  library("eulerr")
  library("RIdeogram")
  library("GenomicRanges")
  library("tictoc")
  library("globaltest")
  library("ebGSEA")
  data("dualmap850kEID")
})

source("meth_functions.R")

```


## Load data

```{r,load1}

bl_mvals <- readRDS(file="bl_mvals.rds")
gu_mvals <- readRDS(file="gu_mvals.rds")

bl_design <- readRDS(file="bl_design_motor.rds")
gu_design <- readRDS(file="gu_design_motor.rds")

bl_lim <- read.csv("limma_blood_motor.csv")
gu_lim <- read.csv("limma_guthrie_motor.csv")


```

## Probe sets

For each gene, extract out the probes.

```{r,probesets}

anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)
promoters <- myann[promoters,]
gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
sets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))
summary(unlist(lapply(sets,length)))

```

## Gene level enrichment for blood

Effect size is estimated with the median t-statistic for each gene.
Significance is estimated using the fry test.
Results are merged for the user.

* PMEA step 196.063 sec elapsed on 8 cores

* Fry step 293.569 sec elapsed on 8 cores

* Fry single core 1205.519 sec

```{r,bl1}

tic()
res_bl <- main(mval=bl_mvals,design=bl_design,sets,cores=detectCores()/2)
res_bl[[1]] <- subset(res_bl[[1]],nprobes!=0) # remove genes with no probes obvs
res_bl_df <- res_bl[[1]] # enrichment results
res_bl_top <- res_bl[[2]] # limma results
toc() ## 502-522 sec elapsed on 8 cores (8.3-8.7 mins)

head(res_bl_df,30) %>%
  kbl(caption="Top DM genes in blood identified with GMEA") %>%
  kable_paper("hover", full_width = F)

par(mfrow=c(2,1))
hist(res_bl_top$t,xlab="probe t-stat",main="blood at assessment") ; grid() ; abline(v=0,lty=1,lwd=3)
hist(res_bl_df$median,xlab="gene median t-stat",main="blood at assessment") ; grid() ; abline(v=0,lty=1,lwd=3)

pdf("hist_bl_motor.pdf")
par(mfrow=c(2,1))
hist(res_bl_top$t,xlab="probe t-stat",main="blood at assessment") ; grid() ; abline(v=0,lty=1,lwd=3)
plot(1)
hist(res_bl_df$median,xlab="gene median t-stat",main="blood at assessment") ; grid() ; abline(v=0,lty=1,lwd=3)
plot(1)
dev.off()
par(mfrow=c(1,1))

```

Now some visualisations.

```{r,bl2}

volcano_plot(res_bl_df)
probe_bias(res_bl_df)
gmea_boxplot(res_bl,sets)

```

## Gene level enrichment for guthrie cards

```{r,gu1}

tic()
res_gu <- main(mval=gu_mvals,design=gu_design,sets,cores=detectCores()/2)
res_gu[[1]] <- subset(res_gu[[1]],nprobes!=0) # remove genes with no probes obvs
res_gu_df <- res_gu[[1]] # enrichment results
res_gu_top <- res_gu[[2]] # limma results
toc() ## 502-522 sec elapsed on 8 cores (8.3-8.7 mins)

head(res_gu_df,30) %>%
  kbl(caption="Top DM genes in Guthrie card identified with GMEA") %>%
  kable_paper("hover", full_width = F)

par(mfrow=c(2,1))
hist(res_gu_top$t,xlab="probe t-stat",main="neonatal Guthrie card") ; grid() ; abline(v=0,lty=1,lwd=3)
hist(res_gu_df$median,xlab="gene median t-stat",main="neonatal Guthrie card") ; grid() ; abline(v=0,lty=1,lwd=3)

pdf("hist_gu_motor.pdf")
par(mfrow=c(2,1))
hist(res_gu_top$t,xlab="probe t-stat",main="neonatal Guthrie card") ; grid() ; abline(v=0,lty=1,lwd=3)
plot(1)
hist(res_gu_df$median,xlab="gene median t-stat",main="neonatal Guthrie card") ; grid() ; abline(v=0,lty=1,lwd=3)
plot(1)
dev.off()
par(mfrow=c(1,1))

```

Now some visualisations.

```{r,gu2}

volcano_plot(res_gu_df)
probe_bias(res_gu_df)
gmea_boxplot(res_gu,sets)

```

## Comparison of blood and guthrie card

Now compare blood and guthrie card

```{r,compare1}

res_bl_df$bl <- res_bl_df$median
res_gu_df$gu <- res_gu_df$median
m1 <- merge(res_bl_df,res_gu_df,by=0)
m2 <- m1[,c("Row.names","bl","gu")]
rownames(m2) <- m2$Row.names
m2$Row.names=NULL

plot(m2$bl,m2$gu,xlab="bl",ylab="gu",
  col = rgb(red = 0.6, green = 0.6, blue = 0.6, alpha = 0.5) ,
  pch=19,cex=0.9)
grid()
abline(v=0,lty=1,lwd=3,col="black") ; abline(h=0,lty=1,lwd=3,col="black")
mtext("aggregated t-statistic for each gene (median)")
#plot(rank(m1$bl),rank(m1$gu),xlab="bl rank",ylab="gu rank",cex=0.6,pch=19) ; grid()
rm2 <- apply(m2,2,rank)
mydiff <- apply(m2,2,function(x) { length(which(x<0)) } )
rm3 <- rm2
rm3[,1] <- rm3[,1] - mydiff[1]
rm3[,2] <- rm3[,2] - mydiff[2]
rnk <- rm3
    palette <- colorRampPalette(c("white", "yellow", "orange", "red",
        "darkred", "black"))
    xmin = min(rnk[, 1])
    xmax = max(rnk[, 1])
    ymin = min(rnk[, 2])
    ymax = max(rnk[, 2])
    k <- MASS::kde2d(rnk[, 1], rnk[, 2])
    X_AXIS = paste("Rank in contrast", colnames(rnk)[1])
    Y_AXIS = paste("Rank in contrast", colnames(rnk)[2])
    filled.contour(k, xlim = c(xmin, xmax), ylim = c(ymin, ymax), 
        color.palette = palette, plot.title = {
            abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
            title(main = "Rank-rank plot of all genes", xlab = X_AXIS, 
            ylab = Y_AXIS)
        })

pdf("genedens_motor.pdf")
par(mfrow=c(2,2))
plot(1)
plot(1)

plot(m2$bl,m2$gu,xlab="bl",ylab="gu",
  col = rgb(red = 0.6, green = 0.6, blue = 0.6, alpha = 0.5) ,
  pch=19,cex=0.9)
grid()
abline(v=0, h=0, lty=2, lwd=2, col="blue")
mtext("aggregated t-statistic for each gene (median)")

    filled.contour(k, xlim = c(xmin, xmax), ylim = c(ymin, ymax), 
        color.palette = palette, plot.title = {
            abline(v = 0, h = 0, lty = 2, lwd = 2, col = "blue")
            title(main = "Rank-rank plot of all genes", xlab = X_AXIS, 
            ylab = Y_AXIS)
        })

dev.off()

```

Now we need a table of genes with top scores.

```{r,topgenes1}

m3 <- m2
m3$med <- apply(rnk,1,median)
mg <- merge(res_gu_df,res_bl_df,by=0)
mg$gu = mg$bl = NULL
mg$med <- m3$med
mg2 <- data.frame(mg,rnk)
mg2$med <- apply(rnk,1,median)
mg2 <- mg2[order(mg2$med),]
mg2 <-subset(mg2,nprobes.x>=5)

head(mg2,20) %>%
  kbl(caption="Genes with coordinated hypomethylation in neonatal Guthrie card and blood at assessment") %>%
  kable_paper("hover", full_width = F)

mg2 <- mg2[order(-mg2$med),]

head(mg2,20) %>%
  kbl(caption="Genes with coordinated hypermethylation in neonatal Guthrie card and blood at assessment") %>%
  kable_paper("hover", full_width = F)

```

## Gene set level enrichment

```{r,genesets1}

genesets <- gmt_import("ReactomePathways.gmt")

cores = 8

```

## Mitch enrichment analysis

Here I'm using the median t-statistic for downstream enrichment.
Pathways are from REACTOME and analysis using mitch.

```{r,mitch_direction,eval=FALSE}

res_bl_df$bl <- res_bl_df$median
res_gu_df$gu <- res_gu_df$median

bl <-  res_bl_df[,"bl",drop=FALSE]
mitch_bl <- mitch_calc(bl,genesets,priority="effect")
head( mitch_bl$enrichment_result,30) %>%
  kbl(caption="BLOOD: Top effect size pathways found with mitch") %>%
  kable_paper("hover", full_width = F)

sig <- subset(mitch_bl$enrichment_result,`p.adjustANOVA`<0.05)
head(sig[order(-abs(sig$s.dist)),],30) %>%
  kbl(caption="BLOOD: Top effect size pathways found with mitch after 1% FDR filtering") %>%
  kable_paper("hover", full_width = F)

gu <-  res_gu_df[,"gu",drop=FALSE]
mitch_gu <- mitch_calc(gu,genesets,priority="effect")
head( mitch_gu$enrichment_result,30) %>%
  kbl(caption="GUTHRIE: Top effect size pathways found with mitch") %>%
  kable_paper("hover", full_width = F)

sig <- subset(mitch_gu$enrichment_result,`p.adjustANOVA`<0.05)
head(sig[order(-abs(sig$s.dist)),],30) %>%
  kbl(caption="GUTHRIE: Top effect size pathways found with mitch after 1% FDR filtering") %>%
  kable_paper("hover", full_width = F)

```

Mitch joint analysis of blood and guthrie card.
As using the median value is apparently the most sensitive, we'll go with that.

```{r,mitch_joint}

m <- merge(res_bl_df,res_gu_df,by=0)
rownames(m) <- m$Row.names
m2 <- m[,c("bl","gu")]
m2[is.na(m2)] <- 0
res <- mitch_calc(m2, genesets, priority="effect")
mitch_report(res,"blgu_mitch.html",overwrite=TRUE)
mitch_plots(res=res,outfile="blgu_mitch_motor.pdf")
sig <- subset(res$enrichment_result,p.adjustMANOVA<0.01)

head(sig[order(-abs(sig$s.dist)),],30) %>%
  kbl(caption="Top pathways found with mitch (effect size after 1% FDR filtering)") %>%
  kable_paper("hover", full_width = F)

```

## Camera enrichment analysis

Camera is designed to control for correlation bias.
Using the median value gives better results than the directional significance or the combined metric.
Blood dataset.
Need to filter by FDR then sort by effect size.

```{r,cam_bl}

stat <- res_bl_df$median
names(stat) <- rownames(res_bl_df)
stat[is.na(stat)] <- 0

tic()
cres <- cameraPR(statistic=stat, index=genesets, use.ranks = FALSE, inter.gene.cor=0.01, sort = TRUE)
cres <- subset(cres,NGenes>4)
cres$FDR <- p.adjust(cres$PValue,method="fdr")
mymedians <- sapply(1:nrow(cres),function(i) {
  myset <- genesets[[which(names(genesets) == rownames(cres)[i])]]
  mystats <- stat[names(stat) %in% myset]
  median(mystats)
})
cres$median <- mymedians
toc() # 1.0 sec

sig <- subset(cres,FDR<0.05)

head(sig[order(-abs(sig$median)),],20) %>%
  kbl(caption = "BLOOD: Top effect size pathways found with CAMERA after 5% FDR filtering") %>%
  kable_paper("hover", full_width = F)

nrow(subset(cres,FDR<0.05 & Direction=="Up"))
nrow(subset(cres,FDR<0.05 & Direction=="Down"))

cres_bl <- cres

```

Guthrie dataset.

```{r,cam_guth}

stat <- res_gu_df$median
names(stat) <- rownames(res_gu_df)
stat[is.na(stat)] <- 0

tic()
cres <- cameraPR(statistic=stat, index=genesets, use.ranks = FALSE, inter.gene.cor=0.01, sort = TRUE)
cres <- subset(cres,NGenes>4)
cres$FDR <- p.adjust(cres$PValue,method="fdr")
mymedians <- sapply(1:nrow(cres),function(i) {
  myset <- genesets[[which(names(genesets) == rownames(cres)[i])]]
  mystats <- stat[names(stat) %in% myset]
  median(mystats)
})
cres$median <- mymedians
toc() # 1.0 sec

sig <- subset(cres,FDR<0.05)

head(sig[order(-abs(sig$median)),],20) %>%
  kbl(caption = "GUTHRIE: Top effect size pathways found with CAMERA after 5% FDR filtering") %>%
  kable_paper("hover", full_width = F)

cres_gu <- cres

```

## Compare CAMERA data

Compare median set score for both contrasts.
In the graph below, the larger points represent larger gene sets.
Red fill means significant in blood contrast.
Blue ring means significant in guthrie contrast.

```{r,camera_compare}

m4 <- merge(cres_bl,cres_gu,by=0)
sigx <- subset(m4,FDR.x<0.05)
sigy <- subset(m4,FDR.y<0.05)

plot(m4$median.x,m4$median.y,col="gray",pch=19,cex=log10(m4$NGenes.x),
  xlab="blood",ylab="Guthrie",main="median t-stat")
grid()
points(sigx$median.x,sigx$median.y,col="red",pch=19,cex=log10(sigx$NGenes.x))
points(sigy$median.x,sigy$median.y,col="blue",pch=1,cex=log10(sigy$NGenes.x))
abline(v=0,h=0,lty=2,lwd=3)

par(mfrow=c(1,1))
pdf("camera_pw_motor.pdf")

plot(m4$median.x,m4$median.y,col="gray",pch=19,cex=log10(m4$NGenes.x),
  xlab="blood",ylab="Guthrie",main="median t-stat")
grid()
points(sigx$median.x,sigx$median.y,col="red",pch=19,cex=log10(sigx$NGenes.x))
points(sigy$median.x,sigy$median.y,col="blue",pch=1,cex=log10(sigy$NGenes.x))
abline(v=0,h=0,lty=2,lwd=3)

dev.off()

subset(m4,FDR.x<0.05 & FDR.y<0.05 ) %>%
  kbl(caption = "Pathways significant in blood (x) and guthrie cards (y) at 5%FDR") %>%
  kable_paper("hover", full_width = F)

```

make scatterplots of the top sets

```{r,topsets}

XMAX=max(res$ranked_profile[,1])
XMIN=min(res$ranked_profile[,1])
YMAX=max(res$ranked_profile[,2])
YMIN=min(res$ranked_profile[,2])
plot(res$detailed_sets[[1]], pch=19,cex=2,col="lightgray",
  xlim=c(XMIN,XMAX),ylim=c(YMIN,YMAX),
  xlab="blood",ylab="Guthrie")
text(res$detailed_sets[[1]],labels=rownames(res$detailed_sets[[1]]))
mtext(names(res$detailed_sets)[[1]])
grid() ; abline(v=0,h=0,lty=2,lwd=3)

pdf("c3_metabotropic_motor.pdf")

plot(res$detailed_sets[[1]], pch=19,cex=2,col="lightgray",
  xlim=c(XMIN,XMAX),ylim=c(YMIN,YMAX), 
  xlab="blood",ylab="Guthrie")
text(res$detailed_sets[[1]],labels=rownames(res$detailed_sets[[1]]))
mtext(names(res$detailed_sets)[[1]])
grid() ; abline(v=0,h=0,lty=2,lwd=3)

dev.off()


```

## Distinguishing ASD from birth

Here I will use MDS plot to cluster samples.

First I will use all probes, then I will focus on a subset of candidate probes.

```{r,mds1}

plot(cmdscale(dist(t(bl_mvals))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "n",main="Blood")
mtext("Numbers indicate motor impairment score")
text(cmdscale(dist(t(bl_mvals))), labels=bl_design[,ncol(bl_design)], )

plot(cmdscale(dist(t(gu_mvals))), xlab="Coordinate 1", ylab="Coordinate 2", 
  type = "n",main="Guthrie card")
mtext("Numbers indicate motor impairment score")
text(cmdscale(dist(t(gu_mvals))), labels=gu_design[,ncol(gu_design)], )

```

There is no apparent clustering by motor impairment score, indicating a subtle effect.

```{r,mds3}

myprobes2 <- head(bl_lim$Name,50)
myprobedat2 <- bl_mvals[which(rownames(bl_mvals) %in% myprobes2),]
plot(cmdscale(dist(t(myprobedat2))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "n",main="Blood")
text(cmdscale(dist(t(myprobedat2))), labels=bl_design[,ncol(bl_design)], )

myprobes2 <- head(gu_lim$Name,50)
myprobedat2 <- gu_mvals[which(rownames(gu_mvals) %in% myprobes2),]
plot(cmdscale(dist(t(myprobedat2))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "n", main="Guthrie")
text(cmdscale(dist(t(myprobedat2))), labels=gu_design[,ncol(gu_design)], )

```

MDS analysis of top probes identified without correction for covariates.

```{r,woc1}

bl_design2 <- bl_design[,c(1,10)]
fit <- lmFit(bl_mvals, bl_design2)
fit <- eBayes(fit)
summary(decideTests(fit))
top <- topTable(fit,coef="motor",num=Inf, sort.by = "P")
head(top)
myprobes3 <- rownames(top)[1:20]

myprobedat3 <- bl_mvals[which(rownames(bl_mvals) %in% myprobes3),]
plot(cmdscale(dist(t(myprobedat3))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "n",main="Blood")
text(cmdscale(dist(t(myprobedat3))), labels=bl_design[,ncol(bl_design)], )

```

```{r,woc2}

gu_design2 <- gu_design[,c(1,12)]
fit <- lmFit(gu_mvals, gu_design2)
fit <- eBayes(fit)
summary(decideTests(fit))
top <- topTable(fit,coef="motor",num=Inf, sort.by = "P")
head(top)
myprobes3 <- rownames(top)[1:20]

myprobedat3 <- gu_mvals[which(rownames(gu_mvals) %in% myprobes3),]
plot(cmdscale(dist(t(myprobedat3))), xlab="Coordinate 1", ylab="Coordinate 2", 
  type = "n",main="Guthrie")
text(cmdscale(dist(t(myprobedat3))), labels=gu_design[,ncol(gu_design)], )

```

Again it was unsuccessful.

## Try ebGSEA

ebGSEA uses globaltest to identify DMGs from infinium data.
Some points to note:

* only 1 sided test - does not distinguish between up and down methylated genes.

* cannot use complex experiment designs.

Therefore I modify how the ebGSEA code uses globaltest.

item mapEIDto850k.lv : A list mapping Entrez Gene ID to 850k probes

```{r,bl_ebgsea}

#bl_gt <- doGT(bl_design,bl_mvals,array="850k",ncores=16)
data.m <- bl_mvals

subsets <- mclapply(mapEIDto850k.lv,intersect,rownames(data.m),mc.cores = 16)
nrep.v <- unlist(lapply(subsets,length))
selG.idx <- which(nrep.v>0)

gt.o <- gt(response=bl_design[,ncol(bl_design)], alternative=t(data.m), model="linear",
  directional = TRUE, standardize = FALSE, permutations = 0,
  subsets=subsets[selG.idx],trace=F)

gt_res <- result(gt.o)
gt_res <- gt_res[order(gt_res$`p-value`),]

```

## Gene set level enrichment - motor impairment

```{r,genesets2}

genesets <- gmt_import("ReactomePathways.gmt")
cores = 8

```

Aggregate mval to genes and examine pathway enrichment in blood and guthries.

```{r,gsmea}

tic()
bl_gsmea <- gsmea(mval=bl_mvals,design=bl_design,probesets=sets,genesets=genesets,cores=8)
toc()
bl_gsmea$res <- subset(bl_gsmea$res,nprobes>=5)
bl_gsmea$res$FDR <- p.adjust(bl_gsmea$res$PValue,method="fdr")
head(bl_gsmea$res,20)  %>%  kbl(caption="Top DM pathways in blood identified with GSMEA") %>%
  kable_paper("hover", full_width = F)


tic()
gu_gsmea <- gsmea(mval=gu_mvals,design=gu_design,probesets=sets,genesets=genesets,cores=8)
toc()
gu_gsmea$res <- subset(gu_gsmea$res,nprobes>=5)
gu_gsmea$res$FDR <- p.adjust(gu_gsmea$res$PValue,method="fdr")
head(gu_gsmea$res,20)  %>%  kbl(caption="Top DM pathways in Guthrie card identified with GSMEA") %>%
  kable_paper("hover", full_width = F)


```

## Session information

For reproducibility

```{r,session}

sessionInfo()

```
