---
title: "Characterisation of the ASD cohort"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: https://github.com/markziemann/asd_meth/cohort_characterisation.Rmd


```{r,libs}

library("kableExtra")
library("magrittr")

```

## Intro

Clinical characterisation of the cohort, shown in table 1 of the manuscript.

There are 3 samplesheets provided.
Separate ones for blood and guthrie and also a combined one where the

  19 ASD_blood_summarysheet.csv
  40 ASD_guthrie_blood_combined_summarysheet.csv
  23 ASD_guthrie_summarysheet.csv

```{r,loaddata}

datg <- read.table("ASD_guthrie_summarysheet.csv",sep=",",header=TRUE)

datb <- read.table("ASD_blood_summarysheet.csv",sep=",",header=TRUE)

dat <- datg

dat$Guthrie_sample <- TRUE

dat$Blood_sample <- datg$SampleName %in% datb$Sample_Name

dat %>%
  kbl(caption="Full sample sheet of the anonymous participant data") %>%
  kable_paper("hover", full_width = FALSE)

```

## Guthrie card arm summary stats

Provide some summary statistics for the Guthrie arm of the project.

```{r,datg}

message("summary stats")
summary(datg)

message("standard deviation")
mysd <- apply(datg,2,sd)
format(mysd,scientific=FALSE)

message("categorical counts")
mytab <- apply(datg,2,table)
mytab

```

## Blood sample arm summary stats

Now for the follow-up blood samples.

```{r,datb}

message("summary stats")
summary(datb)

message("standard deviation")
mysd <- apply(datb,2,sd)
format(mysd,scientific=FALSE)

message("categorical counts")
mytab <- apply(datb,2,table)
mytab

```

## Session information

```{r,session}

sessionInformation()

```
